Concurrent Programming in Harmony

### 2021-01-20 book/paper.tex
Latest

\begin{figure}
\begin{center}
\includegraphics[width=0.9\textwidth]{figures/Up1.png}
\end{center}
\caption{The
\harmonyref{Up.html}{HTML output of running Harmony on \autoref{fig:inc}}.
There are four sections.
The top shows the steps to the problematic state.
Each step shows a thread identifier, the method the thread is executing,
a ``timeline'' containing a block for each instruction executed, the
program counter of the thread, and the values of the shared variables
at those steps.
The middle section shows the line of code that is being executed.
The bottom left shows the bytecode of the program.
The bottom right shows the state of each thread.
Each row in the stack trace of a thread shows
the method that is being evaluated and the values of its method variables.}
\label{fig:inchtml1}
\end{figure}

@@

In our case, Harmony reports a safety violation, in particular it
reports that the assertion and that
\textit{count} has value 1 instead of 2.
Next it reports an execution that failed this assertion.
The program got to the faulty execution in 5 ``steps.''
The output has five columns:
\begin{enumerate}
\item A thread identifier;
\item The main method and argument of the thread;
\item The sequence of program counters of the HVM instructions that the thread executed;
\item The contents of the shared memory.
\end{enumerate}

The first step is initialization.
It sets shared variable \textit{count} to 0 and
shared variable \textit{done} to \texttt{[False, False]}.
If we look at the next three steps, we see that:
\begin{itemize}
\item Thread \texttt{incrementer(0)}
executed instructions 6 through 9, loading the value of
\textit{count} but stopping just before storing 1 into \textit{count};
\item Thread \texttt{incrementer(1)}
executed instructions 6 through 18, storing 1 into
into \textit{count} and storing \texttt{True} into \textit{done}[1];
\item Thread \texttt{incrementer(0)}
continues execution, executing instructions 10 through 28
storing value 1 into \textit{count} (instruction 10),
storing \texttt{True} into \textit{done}[0] (instructions 11 through 15)
finding that \textit{done}[1] is \texttt{True} (instructions 16 through 21),
and finally detecting that the assertion is violated
(instructions 22 through 28).
\end{itemize}

Harmony also generates an HTML file that allows exploring more details
of the execution interactively.
Open the suggested HTML file and you should see something like
\autoref{fig:inchtml1}.
You can go to any point in the execution by clicking on one of the
blocks in the ``execution timelines.''  Using the left and right arrow
keys you can go forwards and backwards one instruction at a time.
With the down and up arrow keys you can go through the execution one
step at a time.
If you hit enter, it will to go to the next line of Harmony code.
If you hover the mouse over a machine instruction, it provides a
brief explanation.

@@

\begin{problems}
\item \autoref{fig:incenter} shows an attempt at trying to fix the code of
    \autoref{fig:inc}.  Run it through Harmony and see what happens.  Based on
    the error output, describe in English what is wrong with the code by describing,
    in broad steps, how running the program can get into a bad state.
\item What if we moved line~6 of \autoref{fig:incenter} to after the \textbf{if}
    statement (between lines~9 and~10)?  Do you think that would work?  Run it through
    Harmony and describe either why it works or why it does not work.
\end{problems}

>>>>

@@

In our case, Harmony reports a safety violation, in particular it
reports that the assertion and that
\textit{count} has value 1 instead of 2.
Next it reports an execution that failed this assertion.
The program got to the faulty execution in 4 ``steps.''
The output has four columns:
\begin{enumerate}
\item A thread identifier;
\item The main method and argument of the thread;
\item The sequence of program counters of the HVM instructions that the thread executed;
\item The contents of the shared memory.
\end{enumerate}

\begin{figure}
\begin{center}
\includegraphics[width=0.9\textwidth]{figures/Up1.pdf}
\end{center}
\caption{The
\harmonyref{Up.html}{HTML output of running Harmony on \autoref{fig:inc}}.
There are four sections.
The top shows the steps to the problematic state.
Each step shows a thread identifier, the method the thread is executing,
a ``timeline'' containing a block for each instruction executed, the
program counter of the thread, and the values of the shared variables
at those steps.
The middle section shows the line of code that is being executed.
The bottom left shows the bytecode of the program.
The bottom right shows the state of each thread.
Each row in the stack trace of a thread shows
the method that is being evaluated and the values of its method variables.}
\label{fig:inchtml1}
\end{figure}

The four steps in the execution are as follows:

\begin{enumerate}
\item Thread \texttt{\_\_init\_\_} (with identifier T0)
sets shared variable \textit{count} to 0 and
shared variable \textit{done} to \texttt{[False, False]}.
\item Thread \texttt{incrementer(0)} (with identifier T1)
executed instructions 6 through 9, loading the value of
\textit{count} but stopping just before storing 1 into \textit{count};
\item Thread \texttt{incrementer(1)} (with identifier T2)
executed instructions 6 through 18, storing 1 into
into \textit{count} and storing \texttt{True} into \textit{done}[1];
\item Thread \texttt{incrementer(0)}
continues execution, executing instructions 10 through 28
storing value 1 into \textit{count} (instruction 10),
storing \texttt{True} into \textit{done}[0] (instructions 11 through 15)
finding that \textit{done}[1] is \texttt{True} (instructions 16 through 21),
and finally detecting that the assertion is violated
(instructions 22 through 28).
\end{enumerate}

Harmony also generates an HTML file that allows exploring more details
of the execution interactively.
Open the suggested HTML file and you should see something like
\autoref{fig:inchtml1}.

In the top right, the HTML file contains the reported issue in red.
Underneath it, a table shows the four steps in the execution.
Instead of listing explicitly the program counters of the executed
instructions, the HTML file contains a list of blocks for each
executed instruction.  We call this the \emph{time line}.
You can click on such a block to see the state of the Harmony
virtual machine just after executing that instruction.
The step that is being executed is highlighted in green.
The table also lists the program counter of the thread at each step,
and the values of the shared variables.

Underneath the table it shows the line of Harmony code that is being
executed in blue.

The bottom left shows
the bytecode of the program being executed.  It has alternating
grey and white sections.  Each section corresponds to a line of
Harmony code.  The instruction that is about to be executed,
if any, is highlighted in red.  (In this case, the state shown
is a failed state and no instruction will be executed next.)
If you hover the mouse over a machine instruction, it provides a
brief explanation of what the instruction does.

The bottom right contains a table with the state of each thread.
The thread that is executing is highlighted in green.
Status information for a thread can include:
\begin{itemize}
\item[] \texttt{runnable}: the thread is runnable but not
currently running.  In Harmony, threads are interleaved and so
at most one thread is actually running;
\item[] \texttt{terminated}: the thread has completed all its
instructions;
\item[] \texttt{failed}: the thread has encountered an error,
such as violating an assertion or divide by zero;
\item[] \texttt{blocked}: the thread cannot make progress until
another thread has updated the shared state;
\item[] \texttt{atomic}: the thread is in \emph{atomic} mode,
not allowing other threads to be scheduled.  This is, for example,
the case when an assertion is being checked;
\item[] \texttt{read-only}: the thread is in \emph{read-only} mode,
not able to modify shared state.  Assertions can execute
arbitrary code including methods, but they are not allowed to
modify the shared state.
\end{itemize}

Next to the status, the table shows the \emph{stack trace} of the
thread.  A stack trace is a list of methods that are being invoked.
In this case, the \texttt{implementer} method does not invoke any
other methods, and so the list is of length 1.
For each entry in the stack trace, it shows the method name and
arguments, as well as the variables of the method.

When you load the HTML file, it shows the state after executing
the last instruction.
As mentioned above, you can go to any point in the execution by
clicking on one of the blocks in the time lines.
There are also various handy keyboard shortcuts:
\begin{itemize}
\item \emph{Right arrow}: go to the next instruction;
\item \emph{Left arrow}: go to the previous instruction;
\item \emph{Down arrow}: go to the next step;
\item \emph{Up arrow}: go to the previous step;
\item \emph{Enter (aka Return)}: go to the next line of Harmony code;
\item \emph{0}: go to the initial state.
\end{itemize}

If you want to see an animation of the entire execution, one instruction
at a time, you can first hit 0 and then hold down the right arrow.
If you want to see it one line of Harmony code at a time, hold down the
enter (aka return) key instead.
If you hold down the down arrow key, the movie will
go by very quickly.

@@

\begin{problems}
\item \autoref{fig:incenter} shows an attempt at trying to fix the code of
    \autoref{fig:inc}.  Run it through Harmony and see what happens.  Based on
    the error output, describe in English what is wrong with the code by describing,
    in broad steps, how running the program can get into a bad state.
\item What if we moved line~5 of \autoref{fig:incenter} to after the \textbf{if}
    statement (between lines~8 and~9)?  Do you think that would work?  Run it through
    Harmony and describe either why it works or why it does not work.
\end{problems}
